Work Fragmentation in Developer Interaction Data

### 2016-08-05 paper_2.tex
Victor's changes

\begin{changedforreviewerlong}
From these research questions, we found the following:
\begin{itemize}
	\item We found an inverse relationship between the number of interruptions and our productivity metrics.
	\item A high proportion of prolonged interruptions is associated with lower productivity metrics, although the effect is less pronounced for high number of interruptions.
	\item The productivity 30 minutes around an interruption is lower than the average.
	\item There are two well-defined patterns of productivity around the interruption, depending on whether the effect afterwards was positive or negative.
	\item There is a peak of productivity after the recovery time that lasts until approximately 10 minutes before the next interruption.
	\item There is a difference between the activities performed after an interruption, depending on the productivity patterns it exhibits (positive or negative).
\end{itemize}
%\RR{For each research question, add a sentence explaining why we look into it}
\end{changedforreviewerlong}

@@

While these studies already show the effect of work fragmentation can be hurtful, they are mostly observational studies over a limited time, and do not address specifically the sub-population of programmers. According to Parnin and Rugaber \cite{PR12}, programmers must perform activities that challenge the memory, like keeping track of refactoring changes, navigate and debug unfamiliar code, learn new APIs, and understand new concepts; when these tasks are interrupted, it can be costly to recover the lost state. Some of these tasks require to invest in program comprehension activities, which can account from 54 to 95\% of the time spent on every working session \cite{MMLK14}.

Program comprehension is especially hard, since programmers need to build a mental model of an existing program; such mental models are often very complex, involving dozens or hundreds of program elements, which is much more than can comfortably fit in memory. Such an activity is especially difficult to interrupt: Parnin and Rugaber describe the various information needs that programmer have in the situations where memory fails \cite{PR12}. Something similar happens in the context of businessman and accountants when building and debugging spreadsheets, as they must keep the context of the information represented in terms of real world problems and therefore they must build complex mental models \cite{KS13}.

@@

\begin{changedforreviewerlong}
\textbf{Selection events. } Mylyn's selection feature is tied to high-level selection events in the Package Explorer. It does not include events such as switching tabs or navigation within and between documents via structured navigation (e.g., goto definition).
\end{changedforreviewerlong}

@@

\begin{changedforreviewerlong}
Mylyn data has another characteristic which is unusual for time series: the time of occurrence of the events is not periodic (see column \textit{StartTime} in Table \ref{tbl:sample_event}). That is to say, events occur with non-equidistant separation gaps between them. However, in a time series, the values must be evenly spaced and chronologically sorted. For this reason, we compressed the size of the multivariate time series, pulling apart all the interruptions as a new time series variable. Then, each time series value represents the interruption duration in this minute (Figure \ref{fig:trans}). This transformation to a time series is similar to the Codealike's focus metric, which creates a time series with the quantification of the user's activity per minute, but without the decay and smoothing functions \cite{CLQ15}.
\end{changedforreviewerlong}

@@

\begin{figure}
	\centering	
	\begin{minipage}[c]{0.30\linewidth}
		\small
		\centering
			\begin{tabular}{| p{2cm} |p{2cm} |} 
				\hline 
				stat & value \\  
				\hline 
				\hline 
				total & 4,284 \\
				min. & 2 min \\
				1st Qu. & 18 min\\
				median & 35 min \\
				mean & 48.91 min \\
				3rd Qu. & 62 min \\
				max. & 535 min\\
				\changedforreviewer{std. dev.} & 102.36 min\\
				\hline

@@

\begin{changedforreviewerlong} 
An important reason to consider the edit ratio is that having an interruption requires doing activities that might increase the execution of selection events, reducing the edit events. For example reading the code, navigating around files and starting a debugging sessions are activities heavily loaded with selections events and are common activities after an interruption, according to observational studies \cite{PR11,PD10}. The edit ratio can tell us when there is a reduction of the propodrtion of edit events due to an increase in selection events.
\end{changedforreviewerlong}

\paragraph{Correlation analysis of the metrics}
%\RR{perhaps expand a bit}
\begin{changedforreviewerlong}
The metrics that characterize the productivity are related to the amount of activity recorded. Thus there is a degree of correlation between them; we study this correlation to determine whether the metrics measure different aspects or not.  %weand could be possible to find a correlation between them, specially with the number of edits and selections per minute, and the edit ratio. 
We did a correlation analysis involving these three metrics (Table \ref{tbl:corr_analysis}). We used Pearson's correltion coefficient, and separated the data between traces with or without interruptions. We found that the Edits/min and Selections/minute metrics have a degree of correlation ($r=0.25$ and $r=0.35$), but that it is relatively low; this is the expected result as they explain two different kinds of activities. We do find that the correlation is higher in traces with interruptions. The same happens between the Selections/minute and Edit ratio metrics with the data of both kind of traces ($r=-0.38$ and $r=-0.32$), this time being a negative correlation which is also an expected value, as the Edit ratio should decrease when the Selections/minute increases.

@@

\end{changedforreviewerlong}

@@

\subsection{Classification of events}
\begin{changedforreviewerlong}
Mylyn events belong to two main categories: edition and selection. The UDC data does not have this distinction, so in order to match the classification of Mylyn we inferred the type of every event manually. We do not base our classification on previous work (to our knowledge there is none) and we are aware that a change or misinterpretation could affect the final results. However, the attributes of the events offer enough information to make an accurate inference in most of the cases.

Note that we reuse the same term of selection in the following text, as it was used in Mylyn data. Strictly speaking, the word "navigation" would be better suited to the variety of events present in the UDC data, but we keep the old term for consistency.
\end{changedforreviewerlong}
%To select the event type, we made a thorough analysis of the different events to classify them where they better fit. In comparison with Mylyn we didn't have a classification by default but the attributes can give us a very good approximation.

@@

\begin{figure}
	\centering	
	\begin{minipage}[c]{0.30\linewidth}
		\small
		\centering
		\begin{tabular}{| p{2cm} |p{2cm} |} 
			\hline 
			stat & value \\  
			\hline 
			\hline 
			total & 15,825 \\
			min. & 2 min \\
			1st Qu. & 11 min\\
			median & 23 min \\
			mean & 35.87 min \\
			3rd Qu. & 47 min \\
			max. & 1,345 min\\
			\changedforreviewer{std. dev.} & 43.95 min\\
			\hline

@@

\begin{changedforreviewerlong}
Beyond visual inspection, we also quantify the statistical and the practical significance of these observations. First, in the results with Mylyn, all the differences observed are statistically significant with very low p-values (see Table~\ref{tbl:p_value}) according to the Mann-Whitney U-test. This is not surprising, given the shape of the boxplots and the size of the samples. It is important to mention that the values from the effect size Tables (\ref{tbl:p_value} and \ref{tbl:p_value_udc}) are obtained by comparing the \textit{none} group with the rest of the groups.
\end{changedforreviewerlong}

@@

   & none & $\leq 3$ & $[4 - 5]$ & $[6 - 10]$ & $\geq 11$  \\  
  \hline
  \multicolumn{6}{c}{\textbf{Edits}} \\
  \hline
  \changedforreviewer{median} & 6.56 & 1 & 0.67 & 0.65 & 0.6	 \\
  \cline{2-6} 
  mean & 6.29 &	2.59 & 1.55 & 1.29 & 0.91  \\ 
  % \cline{3-6} 
  %t.test & $\hookrightarrow$& \multicolumn{4}{c}{$<$ 2.2e-16} \\
   \cline{3-6} 
  U-test (adjusted) & $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
  \cline{3-6} 
  Cliff's delta & $\hookrightarrow$	& \textbf{0.61} & \textbf{0.77} & \textbf{0.80} & \textbf{0.87}    \\
  \hline

  \multicolumn{6}{c}{\textbf{Selections}} \\
  \hline 
  \changedforreviewer{median} & 4.56 & 2.08 & 1.71 & 1.42 & 1.28 \\
  \cline{2-6} 
  mean & 4.73 &	2.93 & 2.26 & 1.79 & 1.54  \\ 
    % \cline{3-6} 
    %t.test & $\hookrightarrow$& 6.258e-08 & 1.623e-12 & 4.87e-16 & $<$ 2.2e-16 \\
     \cline{3-6} 
    U-test (adjusted) & $\hookrightarrow$ & 8.1e-13 & \multicolumn{3}{c}{$<$ 4.4e-16} \\

@@

  \multicolumn{6}{c}{\textbf{Edit ratio}} \\
  \hline 
  \changedforreviewer{median} & 0.56 & 0.36 & 0.30 & 0.32 & 0.33 \\
  \cline{2-6} 
  mean & 0.55 & 0.39 & 0.33 & 0.34 & 0.33 \\ 
  \cline{3-6} 
%    t.test & $\hookrightarrow$& 1.62E-13 & \multicolumn{3}{c}{$<$ 2.2e-16} \\
     \cline{3-6} 
    U-test (adjusted) & $\hookrightarrow$ & 4.2e-14& \multicolumn{3}{c}{$<$ 2.9e-16} \\
    \cline{3-6} 
    Cliff's delta & $\hookrightarrow$ & \textbf{0.47} & \textbf{0.66} & \textbf{0.67} & \textbf{0.73} \\ 
\hline

@@

		& none & $\leq 2$ & $[3 - 4]$ & $[5 - 7]$ & $\geq 7$  \\  
		\hline
		\multicolumn{6}{c}{\textbf{Edits}} \\
		\hline
		\changedforreviewer{median} & 3.58 & 1.5 & 1.23 & 1.15 & 1.18 \\
		\cline{2-6} 
		mean & 3.97 &	2.18 & 1.82 & 1.70 & 1.59 \\ 
		% \cline{3-6} 
		%t.test & $\hookrightarrow$& \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		U-test (adjusted) & $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$	& \textbf{0.42} & \textbf{0.53} & \textbf{0.57} & \textbf{0.63}    \\

		\hline
		\multicolumn{6}{c}{\textbf{Selections}} \\
		\hline 
		\changedforreviewer{median} & 2.54 & 1.66 & 1.5 & 1.5 & 1.4 \\
		\cline{2-6} 
		mean & 3.10 &	2.93 & 1.80 & 1.77 & 1.63  \\ 
		% \cline{3-6} 
		%t.test & $\hookrightarrow$& 6.258e-08 & 1.623e-12 & 4.87e-16 & $<$ 2.2e-16 \\
		\cline{3-6} 
		U-test (adjusted)& $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$	& \textbf{0.29} & \textbf{0.38} & \textbf{0.39} & \textbf{0.49} \\  
		\hline

		\multicolumn{6}{c}{\textbf{Edit ratio}} \\
		\hline 
		\changedforreviewer{median} & 0.57 & 0.49 & 0.45 & 0.43 & 0.44 \\
		\cline{2-6} 
		mean & 0.68 & 0.56 & 0.55 & 0.54 & 0.57 \\ 
		\cline{3-6} 
		%    t.test & $\hookrightarrow$& 1.62E-13 & \multicolumn{3}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		U-test (adjusted) &  $\hookrightarrow$& \multicolumn{4}{c}{$<$ 0.0001} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$ & \textbf{0.19} & \textbf{0.24} & \textbf{0.28} & \textbf{0.26} \\ 
		\hline

@@

We tested the statistical significance of this results and show the results in the Table \ref{tbl:effect_size_dur_mylyn} for Mylyn, and Table \ref{tbl:effect_size_dur_udc} for UDC. For every group according to the number of interruptions, we measured the effect between the low and high groups of sessions, with the Cliff's delta test and U-test. \changedforreviewer{The groups with one or two interruptions have moderate size effects}, in comparison with the other groups with more interruptions that have lower size effects; this matches our observations via visual inspection on a more defined difference between the low and high groups when having a small number of interruptions. Starting from (approximately) three interruptions, the distributions of the metrics are very similar, except in some cases like the edit ratio with seven interruptions in Mylyn, where the median is almost the same but the distribution of the values are different enough to have a medium effect size.

\begin{table}[ht!]
	\tiny
	\renewcommand{\arraystretch}{1.3}
	\caption{\changedforreviewer{Effect size and statistical significance of the relationships between the groups of sessions by proportion of prolonged interruptions and the number of interruptions in Mylyn}}
	\label{tbl:effect_size_dur_mylyn}
	\centering
	\begin{tabular}{l | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm}} 
		Group & 1 & 2 & 3 & 4 & 5 & 6 & 7  \\  
		\hline
		\multicolumn{8}{c}{\textbf{Edits}} \\
		\hline
		U-test (adjusted) & 5.6e-10 & 6.5e-07 & 2.9e-04 & 0.0074 & 0.0021 & 0.32 &  0.39\\
		\hline
		Cliff's delta & \textbf{0.36} & \textbf{0.26} & \textbf{0.23} & \textbf{0.16} & \textbf{0.26} & \textbf{0.08} & \textbf{-0.09} \\
		\hline

@@

\begin{table}[ht!]
	\tiny
	\renewcommand{\arraystretch}{1.3}
	\caption{\changedforreviewer{Effect size and statistical significance of the relationships between the groups of sessions by proportion of prolonged interruptions and the number of interruptions in UDC}}
	\label{tbl:effect_size_dur_udc}
	\centering
	\begin{tabular}{l | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm}} 
		Group & 1 & 2 & 3 & 4 & 5 & 6 & 7  \\  
		\hline
		\multicolumn{8}{c}{\textbf{Edits}} \\
		\hline
		U-test (adjusted) & 1.9e-20 & 4.0e-13 & 0.0011 & 0.027 & 7.2e-08 & 0.055 & 6.5e-06\\
		\hline
		Cliff's delta & \textbf{0.36} & \textbf{0.27} & \textbf{0.08} & \textbf{0.06} & \textbf{0.17} & \textbf{0.07} & \textbf{0.18} \\
		\hline

@@

\begin{changedforreviewerlong}
Table \ref{tbl:local_effect} shows the applied thresholds and the results, accompanied with a typical example of an interruption in each category. This local analysis shows that there are indeed three well-defined groups of interruptions, with the two largest of them having clear effects on the activity in the session. The 45\% of the interruptions are positive in Mylyn, and the 44\% in UDC. The negative interruptions constitute the 44\% in Mylyn and UDC as well. Finally, the neutral interruptions are 11\% of Mylyn and 12\% of UDC interruptions.
\end{changedforreviewerlong}

@@

\begin{changedforreviewerlong}
We can see in the Figure \ref{fig:box_recovery_udc} the evolution of the metrics through
the phases. The edits per minute and edit ratio show a tendency to increase and reach a maximum between the last phase of the recovery time and the peak of the session. Moreover, the fall of the productivity on the last minutes before the next interruptions agrees with the observations on the local analysis of the time around an interruption. From this we can conclude that the first five minutes after an interruption and the moment before the next one are the phases where productivity is more affected, while the phases farther from the interruptions are where the productivity peaks.
\end{changedforreviewerlong}

@@

In RQ3, we saw different patterns of activity around positive and negative interruptions. In particular, we observed that the edits per minute were higher after a positive interruption; on the contrary they were lower after a negative interruption. The effect is even more marked in the \changedforreviewer{first 5 minutes} after the interruption, as can be seen in the Figure \ref{fig:box_recovery_effect_udc}.

@@

\begin{changedforreviewerlong}
A first challenge that we encountered was that, similarly to the earlier research questions, the high number of different commands made it hard to infer broader trends. We hence created an additional classification of the commands we encountered in the activity logs. In a similar manner to the first classification, we followed an iterative process involving two of the authors to classify the events in these categories. Differently to the first classification, we did not limit ourselves to a restricted number of categories: rather, we let the classification emerge from the iterative process. We ended up with 11 categories, as described in Table \ref{tbl:new_events_udc}. We assigned the label and description that better suited the events grouped on that category and the associated type that reflects the majority of those events.
\end{changedforreviewerlong}

@@

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{New category of events. }
	\label{tbl:new_events_udc}
	\centering
	\begin{tabular}{|p{2.5cm}|p{9 cm}| p{2cm} |}
		\hline 
		category & description & \changedforreviewer{type} \\  
		\hline 
		\hline 
		high-nav & High-level navigation through classes, windows and artifacts & navigation \\
		file & Management of files (saving, closing, opening, etc.) & navigation\\
		edit-text & Text edition events & edition \\
		debug & Events executed during debugging sessions & navigation \\
		text-nav & Low-level text navigation & edition \\
		search & Text or code search & navigation \\
		find-replace & Search and replacement of text or code entities & navigation\\
		perspective & Change of perspectives in Eclipse & navigation  \\
		control & Source control events & navigation \\
		refactoring & Refactoring events & edition \\
		clean-build &  Events related to the build system (clean, build or run a project)  & navigation \\
		\hline
	\end{tabular}
\end{table}

@@

\section{Discussion}
\begin{changedforreviewerlong}
The results presented in this work provides of evidence of the effect of work fragmentation observed in interaction data that is in accordance to what observational studies had concluded before. We began by comparing quantities (number and length of interruptions) and working patterns. The related work to this approach suggests that as a consequence to interruptions there is a recovery time that demands extra time to resume the task, so we expected to see a decrease in the metrics as the number of interruptions increases \cite{MGH05,IH07,CHW04}.

@@

Finally, the second group of interruptions seems to be more disruptive as they require more time to reach the average productivity. The activities performed there being related to program comprehension (e.g., navigation, debug, search), are a reflection of the tasks performed during the recovery time as mentioned by Parnin \cite{PR11}.
\end{changedforreviewerlong}

@@

	{\begin{color}{blue}}%

{}%

>>>>

From these research questions, we found the following:
\begin{itemize}
	\item We found an inverse relationship between the number of interruptions and assumed productivity metrics.
	\item A high proportion of prolonged interruptions is associated with lower productivity metrics, although the effect is less pronounced for high number of interruptions.
	\item The productivity 30 minutes around an interruption is lower than the average.
	\item There are two well-defined patterns of productivity around the interruption, depending on whether the effect afterwards was positive or negative.
	\item There is a peak of productivity after the recovery time that lasts until approximately 10 minutes before the next interruption.
	\item There is a difference between the activities performed after an interruption, depending on the productivity patterns it exhibits (positive or negative).
\end{itemize}
%\RR{For each research question, add a sentence explaining why we look into it}

@@

\begin{changedforreviewerlong}
While these studies already show the effect of work fragmentation can be hurtful, they are mostly observational studies over a limited time, and do not analyze in detail the use of tools from where evidence could be taken about work fragmentation. In the case of programmers the use of tools to create software is pervasive; software development environments (IDE) are the contexts where a significant proportion of the work is done and the interaction with them can mark moments of interruption \cite{BK16}.

Similarly, previous studies although explore the nature of activities of information workers, do not characterize them at a level which can be used to understand particular aspects of programming as a professional activity. According to Parnin and Rugaber \cite{PR12}, programmers must perform activities that challenge the memory, like keeping track of refactoring changes, navigate and debug unfamiliar code, learn new APIs, and understand new concepts; when these tasks are interrupted, it can be costly to recover the lost state. Some of these tasks require to invest in program comprehension activities, which can account from 54 to 95\% of the time spent on every working session \cite{MMLK14}.

\end{changedforreviewerlong}
Program comprehension is especially hard, since programmers need to build a mental model of an existing program; such mental models are often very complex, involving dozens or hundreds of program elements, which is much more than can comfortably fit in memory. Such an activity is especially difficult to interrupt: Parnin and Rugaber describe the various information needs that programmer have in the situations where memory fails \cite{PR12}. Something similar happens in the context of businessman and accountants when building and debugging spreadsheets, as they must keep the context of the information represented in terms of real world problems and therefore they must build complex mental models \cite{KS13}.

@@

\textbf{Selection events. } Mylyn's selection feature is tied to high-level selection events in the Package Explorer. It does not include events such as switching tabs or navigation within and between documents via structured navigation (e.g., goto definition).

@@

Mylyn data has another characteristic which is unusual for time series: the time of occurrence of the events is not periodic (see column \textit{StartTime} in Table \ref{tbl:sample_event}). That is to say, events occur with non-equidistant separation gaps between them. However, in a time series, the values must be evenly spaced and chronologically sorted. For this reason, we compressed the size of the multivariate time series, pulling apart all the interruptions as a new time series variable. Then, each time series value represents the interruption duration in this minute (Figure \ref{fig:trans}). This transformation to a time series is similar to the Codealike's focus metric, which creates a time series with the quantification of the user's activity per minute, but without the decay and smoothing functions \cite{CLQ15}.

@@

\begin{figure}
	\centering	
	\begin{minipage}[c]{0.30\linewidth}
		\small
		\centering
			\begin{tabular}{| p{2cm} |p{2cm} |} 
				\hline 
				stat & value \\  
				\hline 
				\hline 
				total & 4,284 \\
				min. & 2 min \\
				1st Qu. & 18 min\\
				median & 35 min \\
				mean & 48.91 min \\
				3rd Qu. & 62 min \\
				max. & 535 min\\
				std. dev. & 102.36 min\\
				\hline

@@

An important reason to consider the edit ratio is that having an interruption requires doing activities that might increase the execution of selection events, reducing the edit events. For example reading the code, navigating around files and starting a debugging sessions are activities heavily loaded with selections events and are common activities after an interruption, according to observational studies \cite{PR11,PD10}. The edit ratio can tell us when there is a reduction of the proportion of edit events due to an increase in selection events.

\paragraph{Correlation analysis of the metrics}
%\RR{perhaps expand a bit}

The metrics that characterize the productivity are related to the amount of activity recorded. Thus there is a degree of correlation between them; we study this correlation to determine whether the metrics measure different aspects or not.  %weand could be possible to find a correlation between them, specially with the number of edits and selections per minute, and the edit ratio. 
We did a correlation analysis involving these three metrics (Table \ref{tbl:corr_analysis}). We used Pearson's correltion coefficient, and separated the data between traces with or without interruptions. We found that the Edits/min and Selections/minute metrics have a degree of correlation ($r=0.25$ and $r=0.35$), but that it is relatively low; this is the expected result as they explain two different kinds of activities. We do find that the correlation is higher in traces with interruptions. The same happens between the Selections/minute and Edit ratio metrics with the data of both kind of traces ($r=-0.38$ and $r=-0.32$), this time being a negative correlation which is also an expected value, as the Edit ratio should decrease when the Selections/minute increases.

@@

@@

\subsection{Classification of events}

Mylyn events belong to two main categories: edition and selection. The UDC data does not have this distinction, so in order to match the classification of Mylyn we inferred the type of every event manually. We do not base our classification on previous work (to our knowledge there is none) and we are aware that a change or misinterpretation could affect the final results. However, the attributes of the events offer enough information to make an accurate inference in most of the cases.

Note that we reuse the same term of selection in the following text, as it was used in Mylyn data. Strictly speaking, the word "navigation" would be better suited to the variety of events present in the UDC data, but we keep the old term for consistency.

%To select the event type, we made a thorough analysis of the different events to classify them where they better fit. In comparison with Mylyn we didn't have a classification by default but the attributes can give us a very good approximation.

@@

\begin{figure}
	\centering	
	\begin{minipage}[c]{0.30\linewidth}
		\small
		\centering
		\begin{tabular}{| p{2cm} |p{2cm} |} 
			\hline 
			stat & value \\  
			\hline 
			\hline 
			total & 15,825 \\
			min. & 2 min \\
			1st Qu. & 11 min\\
			median & 23 min \\
			mean & 35.87 min \\
			3rd Qu. & 47 min \\
			max. & 1,345 min\\
			std. dev. & 43.95 min\\
			\hline

@@

Beyond visual inspection, we also quantify the statistical and the practical significance of these observations. First, in the results with Mylyn, all the differences observed are statistically significant with very low p-values (see Table~\ref{tbl:p_value}) according to the Mann-Whitney U-test. This is not surprising, given the shape of the boxplots and the size of the samples. It is important to mention that the values from the effect size Tables (\ref{tbl:p_value} and \ref{tbl:p_value_udc}) are obtained by comparing the \textit{none} group with the rest of the groups.

@@

   & none & $\leq 3$ & $[4 - 5]$ & $[6 - 10]$ & $\geq 11$  \\  
  \hline
  \multicolumn{6}{c}{\textbf{Edits}} \\
  \hline
  median & 6.56 & 1 & 0.67 & 0.65 & 0.6	 \\
  \cline{2-6} 
  mean & 6.29 &	2.59 & 1.55 & 1.29 & 0.91  \\ 
  % \cline{3-6} 
  %t.test & $\hookrightarrow$& \multicolumn{4}{c}{$<$ 2.2e-16} \\
   \cline{3-6} 
  U-test (adjusted) & $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
  \cline{3-6} 
  Cliff's delta & $\hookrightarrow$	& \textbf{0.61} & \textbf{0.77} & \textbf{0.80} & \textbf{0.87}    \\
  \hline

  \multicolumn{6}{c}{\textbf{Selections}} \\
  \hline 
  median & 4.56 & 2.08 & 1.71 & 1.42 & 1.28 \\
  \cline{2-6} 
  mean & 4.73 &	2.93 & 2.26 & 1.79 & 1.54  \\ 
    % \cline{3-6} 
    %t.test & $\hookrightarrow$& 6.258e-08 & 1.623e-12 & 4.87e-16 & $<$ 2.2e-16 \\
     \cline{3-6} 
    U-test (adjusted) & $\hookrightarrow$ & 8.1e-13 & \multicolumn{3}{c}{$<$ 4.4e-16} \\

@@

  \multicolumn{6}{c}{\textbf{Edit ratio}} \\
  \hline 
  median & 0.56 & 0.36 & 0.30 & 0.32 & 0.33 \\
  \cline{2-6} 
  mean & 0.55 & 0.39 & 0.33 & 0.34 & 0.33 \\ 
  \cline{3-6} 
%    t.test & $\hookrightarrow$& 1.62E-13 & \multicolumn{3}{c}{$<$ 2.2e-16} \\
     \cline{3-6} 
    U-test (adjusted) & $\hookrightarrow$ & 4.2e-14& \multicolumn{3}{c}{$<$ 2.9e-16} \\
    \cline{3-6} 
    Cliff's delta & $\hookrightarrow$ & \textbf{0.47} & \textbf{0.66} & \textbf{0.67} & \textbf{0.73} \\ 
\hline

@@

		& none & $\leq 2$ & $[3 - 4]$ & $[5 - 7]$ & $\geq 7$  \\  
		\hline
		\multicolumn{6}{c}{\textbf{Edits}} \\
		\hline
		median & 3.58 & 1.5 & 1.23 & 1.15 & 1.18 \\
		\cline{2-6} 
		mean & 3.97 &	2.18 & 1.82 & 1.70 & 1.59 \\ 
		% \cline{3-6} 
		%t.test & $\hookrightarrow$& \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		U-test (adjusted) & $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$	& \textbf{0.42} & \textbf{0.53} & \textbf{0.57} & \textbf{0.63}    \\

		\hline
		\multicolumn{6}{c}{\textbf{Selections}} \\
		\hline 
		median & 2.54 & 1.66 & 1.5 & 1.5 & 1.4 \\
		\cline{2-6} 
		mean & 3.10 &	2.93 & 1.80 & 1.77 & 1.63  \\ 
		% \cline{3-6} 
		%t.test & $\hookrightarrow$& 6.258e-08 & 1.623e-12 & 4.87e-16 & $<$ 2.2e-16 \\
		\cline{3-6} 
		U-test (adjusted)& $\hookrightarrow$ & \multicolumn{4}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$	& \textbf{0.29} & \textbf{0.38} & \textbf{0.39} & \textbf{0.49} \\  
		\hline

		\multicolumn{6}{c}{\textbf{Edit ratio}} \\
		\hline 
		median & 0.57 & 0.49 & 0.45 & 0.43 & 0.44 \\
		\cline{2-6} 
		mean & 0.68 & 0.56 & 0.55 & 0.54 & 0.57 \\ 
		\cline{3-6} 
		%    t.test & $\hookrightarrow$& 1.62E-13 & \multicolumn{3}{c}{$<$ 2.2e-16} \\
		\cline{3-6} 
		U-test (adjusted) &  $\hookrightarrow$& \multicolumn{4}{c}{$<$ 0.0001} \\
		\cline{3-6} 
		Cliff's delta & $\hookrightarrow$ & \textbf{0.19} & \textbf{0.24} & \textbf{0.28} & \textbf{0.26} \\ 
		\hline

@@

We tested the statistical significance of this results and show the results in the Table \ref{tbl:effect_size_dur_mylyn} for Mylyn, and Table \ref{tbl:effect_size_dur_udc} for UDC. For every group according to the number of interruptions, we measured the effect between the low and high groups of sessions, with the Cliff's delta test and U-test. The groups with one or two interruptions have moderate size effects, in comparison with the other groups with more interruptions that have lower size effects; this matches our observations via visual inspection on a more defined difference between the low and high groups when having a small number of interruptions. Starting from (approximately) three interruptions, the distributions of the metrics are very similar, except in some cases like the edit ratio with seven interruptions in Mylyn, where the median is almost the same but the distribution of the values are different enough to have a medium effect size.

\begin{table}[ht!]
	\tiny
	\renewcommand{\arraystretch}{1.3}
	\caption{Effect size and statistical significance of the relationships between the groups of sessions by proportion of prolonged interruptions and the number of interruptions in Mylyn}
	\label{tbl:effect_size_dur_mylyn}
	\centering
	\begin{tabular}{l | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm}} 
		Group & 1 & 2 & 3 & 4 & 5 & 6 & 7  \\  
		\hline
		\multicolumn{8}{c}{\textbf{Edits}} \\
		\hline
		U-test (adjusted) & 5.6e-10 & 6.5e-07 & 2.9e-04 & 0.0074 & 0.0021 & 0.32 &  0.39\\
		\hline
		Cliff's delta & \textbf{0.36} & \textbf{0.26} & \textbf{0.23} & \textbf{0.16} & \textbf{0.26} & \textbf{0.08} & \textbf{-0.09} \\
		\hline

@@

\begin{table}[ht!]
	\tiny
	\renewcommand{\arraystretch}{1.3}
	\caption{Effect size and statistical significance of the relationships between the groups of sessions by proportion of prolonged interruptions and the number of interruptions in UDC}
	\label{tbl:effect_size_dur_udc}
	\centering
	\begin{tabular}{l | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm} | p{1cm}} 
		Group & 1 & 2 & 3 & 4 & 5 & 6 & 7  \\  
		\hline
		\multicolumn{8}{c}{\textbf{Edits}} \\
		\hline
		U-test (adjusted) & 1.9e-20 & 4.0e-13 & 0.0011 & 0.027 & 7.2e-08 & 0.055 & 6.5e-06\\
		\hline
		Cliff's delta & \textbf{0.36} & \textbf{0.27} & \textbf{0.08} & \textbf{0.06} & \textbf{0.17} & \textbf{0.07} & \textbf{0.18} \\
		\hline

@@

Table \ref{tbl:local_effect} shows the applied thresholds and the results, accompanied with a typical example of an interruption in each category. This local analysis shows that there are indeed three well-defined groups of interruptions, with the two largest of them having clear effects on the activity in the session. The 45\% of the interruptions are positive in Mylyn, and the 44\% in UDC. The negative interruptions constitute the 44\% in Mylyn and UDC as well. Finally, the neutral interruptions are 11\% of Mylyn and 12\% of UDC interruptions.

@@

We can see in the Figure \ref{fig:box_recovery_udc} the evolution of the metrics through
the phases. The edits per minute and edit ratio show a tendency to increase and reach a maximum between the last phase of the recovery time and the peak of the session. Moreover, the fall of the productivity on the last minutes before the next interruptions agrees with the observations on the local analysis of the time around an interruption. From this we can conclude that the first five minutes after an interruption and the moment before the next one are the phases where productivity is more affected, while the phases farther from the interruptions are where the productivity peaks.

@@

In RQ3, we saw different patterns of activity around positive and negative interruptions. In particular, we observed that the edits per minute were higher after a positive interruption; on the contrary they were lower after a negative interruption. The effect is even more marked in the first 5 minutes after the interruption, as can be seen in the Figure \ref{fig:box_recovery_effect_udc}.

@@

A first challenge that we encountered was that, similarly to the earlier research questions, the high number of different commands made it hard to infer broader trends. We hence created an additional classification of the commands we encountered in the activity logs. In a similar manner to the first classification, we followed an iterative process involving two of the authors to classify the events in these categories. Differently to the first classification, we did not limit ourselves to a restricted number of categories: rather, we let the classification emerge from the iterative process. We ended up with 11 categories, as described in Table \ref{tbl:new_events_udc}. We assigned the label and description that better suited the events grouped on that category and the associated type that reflects the majority of those events.

@@

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{New category of events. }
	\label{tbl:new_events_udc}
	\centering
	\begin{tabular}{|p{2.5cm}|p{9 cm}| p{2cm} |}
		\hline 
		category & description & type \\  
		\hline 
		\hline 
		high-nav & High-level navigation through classes, windows and artifacts & navigation \\
		file & Management of files (saving, closing, opening, etc.) & navigation\\
		edit-text & Text edition events & edition \\
		debug & Events executed during debugging sessions & navigation \\
		text-nav & Low-level text navigation & edition \\
		search & Text or code search & navigation \\
		find-replace & Search and replacement of text or code entities & navigation\\
		perspective & Change of perspectives in Eclipse & navigation  \\
		control & Source control events & navigation \\
		refactoring & Refactoring events & edition \\
		clean-build &  Events related to the build system (clean, build or run a project)  & navigation \\
		\hline
	\end{tabular}
\end{table}

@@

\section{Discussion}

The results presented in this work provides of evidence of the effect of work fragmentation observed in interaction data that is in accordance to what observational studies had concluded before. We began by comparing quantities (number and length of interruptions) and working patterns. The related work to this approach suggests that as a consequence to interruptions there is a recovery time that demands extra time to resume the task, so we expected to see a decrease in the metrics as the number of interruptions increases \cite{MGH05,IH07,CHW04}.

@@

Finally, the second group of interruptions seems to be more disruptive as they require more time to reach the average productivity. The activities performed there being related to program comprehension (e.g., navigation, debug, search), are a reflection of the tasks performed during the recovery time as mentioned by Parnin \cite{PR11}.

@@
